Efficient entropy stable Gauss collocation methods

### 2019-03-07 docs/ESDG_GQ/arxiv/esdg_gq.tex

%\documentclass[preprint,10pt]{elsarticle}
%\documentclass[preprint,10pt]{amsart}
%\documentclass[review, 10pt]{siamart0216}
%\documentclass{siamart0216}
\documentclass[review,onefignum,onetabnum,final]{siamart171218}

@@

Let $\bm{D}$ denote the nodal differentiation matrix, and let $\bm{V}_f$ denote the \note{$2\times (N+1)$} matrix which interpolates polynomials at Gauss nodes to values at endpoints.  These two matrices are defined entrywise as
\[
\bm{D}_{ij} = \LRl{\pd{\ell_j}{x}}_{x = x_i}, \qquad \LRp{\bm{V}_f}_{1i} = \ell_i(-1), \qquad  \LRp{\bm{V}_f}_{2i} = \ell_i(1).
\]
We also introduce the diagonal matrix of quadrature weights $\bm{W}_{ij} = \delta_{ij} w_i$, as well as the one-dimensional mass matrix $\bm{M}$ whose entries are $L^2$ inner products of basis functions.  We assume that these inner products are computed using the quadrature rule $(x_i, w_i)$ at which the solution is collocated.  Under such an assumption, the mass matrix is diagonal with entries equal to the quadrature weights
\[
\bm{M}_{ij} = \int_{-1}^1 \ell_i(x)\ell_j(x) \approx \sum_{k=1}^{N+1} \ell_i(x_k)\ell_j(x_k) w_k = \delta_{ij} w_i = \bm{W}_{ij}.
\]
Since $\bm{M} = \bm{W}$ under a collocation assumption, we utilize $\bm{W}$ for the remainder of this work to emphasize that the mass matrix is diagonal and related to the quadrature weights $w_i$.  The treatment of non-diagonal mass matrices is covered in \cite{chan2017discretely, chan2018discretely}.

@@

Let $f(x), g(x)$ denote two functions, and let $\bm{f}, \bm{g}$ denote the values of $f,g$ at interior nodal points.  We also define vectors $\bm{f}_N, \bm{g}_N$ denoting the values of $f,g$ at both interior and boundary points 
\begin{equation}
\bm{f}_N = \begin{bmatrix}
f\LRp{x_1}\\
\vdots\\
f\LRp{x_{N+1}}\\
f(-1)\\
f(1)
\end{bmatrix} = \begin{bmatrix}
\bm{f} \\
\bm{f}_f
\end{bmatrix}, \qquad
\bm{g}_N = \begin{bmatrix}
g\LRp{x_1}\\
\vdots\\
g\LRp{x_{N+1}}\\
g(-1)\\
g(1)
\end{bmatrix} = \begin{bmatrix}
\bm{g} \\
\bm{g}_f
\end{bmatrix}.
\label{eq:fg}
\end{equation}
Then, a polynomial approximation to $f\pd{g}{x}$ can be computed using $\bm{Q}_N$.  Let $\bm{u}$ denote the nodal values of the polynomial $u(x) \approx f\pd{g}{x}$.  These coefficients are computed via
\begin{equation}
\bm{W}\bm{u} = \begin{bmatrix}
\bm{I}\\
\bm{V}_f
\end{bmatrix}^T \diag{\bm{f}_N} \bm{Q}_N \bm{g}_N.
\label{eq:qn}
\end{equation}
The approximation (\ref{eq:qn}) can be rewritten in ``strong'' form as follows
\begin{align*}
\bm{u} = \bm{W}^{-1}\begin{bmatrix}
\bm{I}\\
\bm{V}_f
\end{bmatrix}^T& \diag{\bm{f}_N} \bm{Q}_N \bm{g}_N \\
= \diag{\bm{f}}\bm{D}\bm{g} &+ \frac{1}{2}\diag{\bm{f}} \bm{W}^{-1} \bm{V}_f^T \bm{B}\LRp{\bm{g}_f - \bm{V}_f\bm{g}} \\
&+ \frac{1}{2}\bm{W}^{-1} \bm{V}_f^T\bm{B}\diag{\bm{f}_f} \LRp{\bm{g}_f - \bm{V}_f\bm{g}},
\end{align*}
where we have used the fact that diagonal matrices commute to simplify expressions.  The decoupled SBP operator $\bm{Q}_N$ can thus be interpreted as adding boundary corrections to the GSBP operator $\bm{D}$ in a skew-symmetric fashion.  \note{More specifically, the expression (\ref{eq:qn}) corresponds to a quadrature approximation of the following variational approximation of the derivative \cite{chan2017discretely}: find $u\in P^N([-1,1])$ such that},
\[
\note{\int_{-1}^1u({x})v({x})  = \int_{-1}^1 {I_N f\pd{I_Ng}{x}v} + \LRl{(g-I_Ng)\frac{\LRp{fv + I_N(fv)}}{2}}_{-1}^1,} \qquad \forall v\in P^N([-1,1]),
\]
\note{where $I_N$ denotes the degree $N$ polynomial approximation at the $(N+1)$ Gauss points.}

@@

Finally, we define $\bm{F}_S$ as the matrix of evaluations of the two-point flux $\bm{f}_S$ at combinations of values of $\tilde{\bm{u}}$
\[
\LRp{\bm{F}_S}_{ij} \myeq \bm{f}_S\LRp{\tilde{\bm{u}}_i,\tilde{\bm{u}}_j}, \quad 1\leq i,j \leq 2(N+3).
\]
Note that, due to the consistency of $\bm{f}_S$, the diagonal of $\bm{F}_S$ reduces to flux evaluations
\begin{equation}
\label{eq:diagFS}
\LRp{\bm{F}_S}_{ii} = \bm{f}_S\LRp{\tilde{\bm{u}}_i,\tilde{\bm{u}}_i} = \bm{f}\LRp{\tilde{\bm{u}}_i}.  
\end{equation}
We can now construct a semi-discretely entropy conservative formulation based on decoupled SBP operators:
\begin{theorem}
\label{thm:consentropy}
Let $\bm{Q}_h$ be defined by (\ref{eq:decoupledsbp2elem}), and let $\bm{u}_h$ denote the two-element solution of the following formulation:
\begin{gather}
\label{eq:form}
\bm{W}_h \td{}{t} \bm{u}_h + \note{2} \bm{I}_h^T\LRp{\bm{Q}_h \circ \bm{F}_S}\bm{1} = 0.
\end{gather}
Then, $\bm{u}_h$ satisfies a semi-discrete conservation of entropy
\begin{gather*}
\bm{1}^T\bm{W}_h\td{S(\bm{u}_h)}{t} + \bm{1}^T\bm{B}\LRp{\bm{v}_f^T\bm{f}(\tilde{\bm{u}}_f) - \psi\LRp{\tilde{\bm{u}}_f}} = 0\\
\bm{v}_f \myeq \begin{bmatrix}
\bm{t}_L^T \bm{v}\LRp{\bm{u}_N^1}\\
\bm{t}_R^T \bm{v}\LRp{\bm{u}_N^2}
\end{bmatrix}, \qquad \tilde{\bm{u}}_f \myeq \bm{u}(\bm{v}_f).
\end{gather*}
%Let $\circ$ denote the Hadamard product, and let $\tilde{\bm{u}}_f^+$ denote values of the entropy-projected conservative variables on neighboring elements.  Then, the formulation
%\begin{equation}
%\label{eq:form}
%\bm{W}\td{\bm{u}_N}{t} + 2 \begin{bmatrix}
%\bm{I}\\
%\bm{V}_f
%\end{bmatrix}^T\LRp{\bm{Q}_N \circ \bm{F}_S}\bm{1} + \bm{V}_f^T \bm{B} \LRp{\bm{f}_S\LRp{\tilde{\bm{u}}_f^+,\tilde{\bm{u}}_f} - \bm{f}(\bm{u})} = 0
%\end{equation}
%satisfies the quadrature form of the semi-discrete conservation of entropy
%\[
%\bm{1}^T\bm{W}\td{S(\bm{u}_N)}{t} + \bm{1}^T\bm{B} \LRp{\bm{v}_f^T\bm{f}(\tilde{\bm{u}}_f) - \psi_i\LRp{\tilde{\bm{u}}_f}} = 0.
%\]
\end{theorem}
\begin{proof}
The proof results from testing with $\bm{v}_h^T$.  Since $\bm{v}(\bm{u}) = \pd{S(\bm{u})}{\bm{u}}$ and $\bm{W}_h$ is diagonal, the time term yields
\[
\bm{v}_h^T\td{}{t}\bm{W}_h\bm{u}_h = \bm{1}^T \td{}{t}\bm{W}_h\diag{\bm{v}_h}\bm{u}_h = \bm{1}^T\td{}{t}\bm{W}_h S(\bm{u}_h).
\]
The spatial term can be manipulated as follows
\begin{align*}
2\bm{v}_h^T \bm{I}_h^T\LRp{\bm{Q}_h \circ \bm{F}_S}\bm{1} &= 2\LRp{\bm{I}_h\bm{v}_h}^T \LRp{\LRp{\bm{S}_h \circ \bm{F}_S}\bm{1} + \frac{1}{2}\LRp{\bm{B}_h \circ \bm{F}_S}\bm{1}}\\
&= \tilde{\bm{v}}^T\LRp{\bm{B}_h \circ \bm{F}_S}\bm{1} + \tilde{\bm{v}}^T \LRp{\bm{S}_h \circ \bm{F}_S}\bm{1} - \bm{1}^T \LRp{\bm{S}_h \circ \bm{F}_S}\tilde{\bm{v}},
\end{align*}
where we have used the skew-symmetry of $\bm{S}_h$ in the last step.  The boundary term reduces to 
\[
\tilde{\bm{v}}^T\LRp{\bm{B}_h \circ \bm{F}_S}\bm{1} = \bm{1}^T\bm{B}_h\tilde{\bm{v}}^T\bm{f}\LRp{\tilde{\bm{u}}}= \bm{1}^T\bm{B}\tilde{\bm{v}}_f^T\bm{f}\LRp{\tilde{\bm{u}_f}},
\]
where we have used (\ref{eq:diagFS}) and the fact that $\bm{B}_h$ is diagonal.  Here, $\tilde{\bm{v}}^T\bm{f}\LRp{\tilde{\bm{u}}}$, $\tilde{\bm{v}}_f^T\bm{f}\LRp{\tilde{\bm{u}}_f}$ denote vectors whose entries are the vector inner products of components of $\tilde{\bm{v}}$ and $\bm{f}\LRp{\tilde{\bm{u}}}$ at volume and face points, respectively.

@@

\note{Extending the two-element case to multiple elements can be done by defining analogous ``global'' differentiation matrices.  Suppose that there are now three elements.  Then, the matrices $\bm{S}_h, \bm{B}_h$ can be defined as}
\begin{gather*}
\note{\bm{S}_h \myeq \LRs{\begin{array}{cc|cc|cc}
\bm{S} &  \frac{1}{2}\bm{V}_f^T\bm{B} & &&& \\
 - \frac{1}{2}\bm{B}\bm{V}_f &  & & \frac{1}{2}\bm{e}_R\bm{e}_L^T &&\\ [.2em]
 \hline\\[-1.0em]
&& \bm{S} &  \frac{1}{2}\bm{V}_f^T\bm{B}&& \\
& -\frac{1}{2}\bm{e}_L\bm{e}_R^T &  - \frac{1}{2}\bm{B}\bm{V}_f &&\frac{1}{2}\bm{e}_R\bm{e}_L^T&\\
 \hline\\[-1.0em]
&&&& \bm{S} &  \frac{1}{2}\bm{V}_f^T\bm{B} \\
&&& -\frac{1}{2}\bm{e}_L\bm{e}_R^T &  - \frac{1}{2}\bm{B}\bm{V}_f &\\
\end{array}}}
\\
\note{\bm{B}_h \myeq \LRs{\begin{array}{cc|c|cc}
&  & && \\
 & -\bm{e}_L\bm{e}_L^T & && \\
 \hline
&&  \bm{0} & & \\
 \hline
&&&&    \\
&&&  & \bm{e}_R\bm{e}_R^T\\
\end{array}}, \qquad \bm{I}_h = \begin{bmatrix}
\bm{I} &&\\
\bm{V}_f &&\\
&\bm{I}&\\
&\bm{V}_f&\\
&&\bm{I}\\
&&\bm{V}_f\\
\end{bmatrix}.}
\end{gather*}
\note{These global operators again satisfy the SBP property $\bm{Q}_h + \bm{Q}_h^T = \bm{B}_h$, where $\bm{Q}_h = \bm{S}_h + \frac{1}{2}\bm{B}_h$.  Moreover, $\bm{Q}_h$ can be used to produce high order accurate approximations of derivatives, and can be used to construct entropy conservative schemes as in Theorem~\ref{thm:consentropy}.  The extension to a general number of elements is straightforward. }

It is possible to convert the semi-discrete entropy equality in Theorem~\ref{thm:consentropy} to a semi-discrete entropy inequality by adding appropriate interface dissipation terms, such as Lax-Friedrichs or matrix dissipation \cite{winters2017uniquely}.  We note that these terms must be computed in terms of $\tilde{\bm{u}}_f$ in order to ensure a discrete dissipation of entropy \cite{chen2017entropy, chan2017discretely}.  \note{Boundary conditions can also be incorporated into the formulation (\ref{eq:form}) in a weak fashion.  Let $\bm{u}_L, \bm{u}_R$ denote the values of the solution at the left and right domain boundaries of a 1D domain.  Using the SBP property, we can modify (\ref{eq:form}) to }
\[
\note{\bm{W}_h\td{\bm{u}_h}{t} + 2\bm{I}_h^T \LRp{\bm{S}_h\circ \bm{F}_S}\bm{1} + \bm{I}_h^T\bm{B}_h \begin{bmatrix}\bm{0}\\ \bm{f}(\bm{u}_L) \\ \vdots \\ \bm{0}\\ \bm{f}(\bm{u}_R)\end{bmatrix}}= 0,
\]
\note{where we have used the consistency of $\bm{f}_S$ and the fact that $\bm{B}_h$ is diagonal to simplify the boundary term.  We can now replace $\bm{f}(\bm{u}_L), \bm{f}(\bm{u}_R)$ with numerical fluxes $\bm{f}^*_L, \bm{f}^*_R$ as follows:}
\begin{equation}
\note{\bm{W}_h\td{\bm{u}_h}{t} + 2\bm{I}_h^T \LRp{\bm{S}_h\circ \bm{F}_S}\bm{1} + \bm{I}_h^T \bm{B}_h \begin{bmatrix}\bm{0}\\ \bm{f}^*_L \\ \vdots \\ \bm{0}\\ \bm{f}^*_R\end{bmatrix}}= 0.
\label{eq:bform}
\end{equation}
\note{If the boundary numerical flux satisfies the entropy stability conditions}
\[
\note{\psi_L - \bm{v}_L^T\bm{f}^*_{L} \leq 0, \qquad {\psi}_R - \bm{v}_R^T\bm{f}^*_{R} \leq 0}
\]
\note{then the resulting scheme satisfies a global entropy inequality \cite{chen2017entropy}.  }

\note{Assuming that the ODE system (\ref{eq:form}) exactly integrated in time and that entropy dissipative numerical fluxes and boundary conditions are used, the solution will satisfy a discrete entropy inequality (\ref{eq:weakentropyineq}).  In practice, the system (\ref{eq:form}) is solved using an ODE time-stepper.  For an explicit time-stepper, then all that is necessary is to invert the diagonal matrix $\bm{W}_h$ and evaluate the spatial terms in (\ref{eq:bform}).  We note that the conservation or dissipation of entropy is guaranteed up to time-stepper accuracy.  In practice, we observe that entropy is dissipated for all problems considered, despite the fact that the proof does not hold at the fully discrete level.}

@@

\section{Numerical results}
\label{sec:3}

\note{In this section, we present numerical examples for} the compressible Euler equations, which are given in $d$ dimensions as follows:
\begin{align*}
\pd{\rho}{t} &+ \note{\sum_{j=1}^d \pd{\LRp{\rho {u}_j}}{x_j}} = 0,\\
\note{\pd{\rho {u}_i}{t}} &+ \note{\sum_{j=1}^d \pd{\LRp{\rho u_iu_j + p\delta_{ij} }}{x_j}} = 0, \qquad i = 1,\ldots,d\\
\pd{E}{t} &+ \note{\sum_{j=1}^d \pd{\LRp{{u}_j(E+p)}}{x_j}} = 0.\nonumber
\end{align*}
Here, $\rho$ is density, \note{$\LRp{{u}_1, \ldots,{u}_d}$} is the vector of velocities, and $E$ is the total energy.  The pressure $p$ and specific internal energy $\rho e$ are given by 
\[
\note{p = (\gamma-1)\LRp{E - \frac{1}{2}\rho \LRp{\sum_{j=1}^d u_j^2}}}, \qquad \note{\rho e = E - \frac{1}{2}\rho \LRp{\sum_{j=1}^d u_j^2}}. %\qquad \LRb{\bm{u}}^2 = \LRp{\sum_{j=1}^d u_j^2}.
\]

@@

v_{1+ i}&= \frac{\rho \note{{u}_i}}{\rho e}, \qquad i = 1,\ldots, d,\\

@@

 \rho \note{u_i} &= (\rho e) v_{1+i}, \qquad i = 1,\ldots,d\\

@@

\note{Let $f$ denote some piecewise polynomial function, and let $f^+$ denote the exterior value of $f$ across an element face.  We define the average and logarithmic averages as follows:}
\[
\note{\avg{f} = \frac{f^+ + f}{2}}, \qquad \note{\avg{f}^{\log} = \frac{f+ - f}{\log\LRp{f^+}-\log\LRp{f}}}.
\]
\note{The average and logarithmic average are assumed to act component-wise on vector-valued functions.  We evaluate the logarithmic average using the numerically stable approach of \cite{ismail2009affordable}.}

@@

&f^1_{1,S}(\bm{u}_L,\bm{u}_R) = \avg{\rho}^{\log} \avg{u_1},& &f^1_{2,S}(\bm{u}_L,\bm{u}_R) = \avg{\rho}^{\log} \avg{u_2},&\\
&f^2_{1,S}(\bm{u}_L,\bm{u}_R) = f^1_{1,S} \avg{u_1} + p_{\rm avg},&  &f^2_{2,S}(\bm{u}_L,\bm{u}_R) = f^1_{2,S} \avg{u_1},&\nonumber\\
&f^3_{1,S}(\bm{u}_L,\bm{u}_R) = f^2_{2,S},& &f^3_{2,S}(\bm{u}_L,\bm{u}_R) = f^1_{2,S} \avg{u_2} + p_{\rm avg},&\nonumber\\
&f^4_{1,S}(\bm{u}_L,\bm{u}_R) = \LRp{E_{\rm avg} + p_{\rm avg}}\avg{u_1},& &f^4_{2,S}(\bm{u}_L,\bm{u}_R) = \LRp{E_{\rm avg} + p_{\rm avg} }\avg{u_2},& \nonumber

@@

where we have introduced the auxiliary quantities 

@@

\note{\beta = \frac{\rho}{2p}}, \qquad p_{\rm avg} = \frac{\avg{\rho}}{2\avg{\beta}}, \qquad \qquad E_{\rm avg} = \frac{\avg{\rho}^{\log}}{2\avg{\beta}^{\log}\LRp{\gamma -1}}   + \frac{\note{u_{\rm avg}^2}}{2}, \label{eq:fluxaux} \\
\note{u^2_{\rm avg}} = 2(\avg{u_1}^2 + \avg{u_1}^2) - \LRp{\avg{u_1^2} +\avg{u_2^2}} \nonumber.  

@@

\avg{\rho}^{\log}\avg{u_1}\\
\avg{\rho}^{\log}\avg{u_1}^2 + p_{\rm avg}\\
\avg{\rho}^{\log}\avg{u_1}\avg{u_2}\\
\avg{\rho}^{\log}\avg{u_1}\avg{u_3}\\
(E_{\rm avg}+ p_{\rm avg})\avg{u_1}\\

@@

\avg{\rho}^{\log}\avg{u_2}\\
\avg{\rho}^{\log}\avg{u_1}\avg{u_2}\\
\avg{\rho}^{\log}\avg{u_2}^2 + p_{\rm avg}\\
\avg{\rho}^{\log}\avg{u_2}\avg{u_3}\\
(E_{\rm avg}+ p_{\rm avg})\avg{u_2}\\

@@

\avg{\rho}^{\log}\avg{u_3}\\
\avg{\rho}^{\log}\avg{u_1}\avg{u_3}\\
\avg{\rho}^{\log}\avg{u_2}\avg{u_3}\\
\avg{\rho}^{\log}\avg{u_3}^2 + p_{\rm avg}\\
(E_{\rm avg}+ p_{\rm avg})\avg{u_3}\\

@@

%\note{where $u,v,w$ denote the $x,y,z$ velocities, respectively, and} 
where the auxiliary quantities are defined as
\note{\beta = \frac{\rho}{2p}}, \qquad p_{\rm avg} = \frac{\avg{\rho}}{2\avg{\beta}}, \qquad E_{\rm avg} = \frac{\avg{\rho}^{\log}}{2(\gamma-1)\avg{\beta}^{\log}} + \frac{1}{2}\avg{\rho}^{\log}\note{{u}^2_{\rm avg}}, \\
\note{u^2_{\rm avg}} = 2(\avg{u_1}^2 + \avg{u_2}^2 + \avg{u_3}^2) - \LRp{\avg{u_1^2} +\avg{u_2^2} + \avg{u_3^2}}.\nonumber

>>>>

%\documentclass[preprint,10pt]{elsarticle}
%\documentclass[preprint,10pt]{amsart}
%\documentclass[review, 10pt]{siamart0216}
%\documentclass{siamart0216}
\documentclass[onefignum,onetabnum,final]{siamart171218}

@@

Let $\bm{D}$ denote the nodal differentiation matrix, and let $\bm{V}_f$ denote the $(N+1)\times 2$ matrix which interpolates polynomials at Gauss nodes to values at endpoints.  These two matrices are defined entrywise as
\[
\bm{D}_{ij} = \LRl{\pd{\ell_j}{x}}_{x = x_i}, \qquad \LRp{\bm{V}_f}_{1i} = \ell_i(-1), \qquad  \LRp{\bm{V}_f}_{2i} = \ell_i(1).
\]
We also introduce the diagonal matrix of quadrature weights $\bm{W}_{ij} = \delta_{ij} w_i$, as well as the one-dimensional mass matrix $\bm{M}$ whose entries are $L^2$ inner products of basis functions.  We assume that these inner products are computed using the quadrature rule $(x_i, w_i)$ at which the solution is collocated.  Under such an assumption, the mass matrix is diagonal with entries equal to the quadrature weights
\[
\bm{M}_{ij} = \int_{-1}^1 \ell_i(x)\ell_j(x) \approx \sum_{k=1}^{N+1} \ell_i(x_k)\ell_j(x_k) w_k = \delta_{ij} w_i = \bm{W}_{ij}.
\]
Since $\bm{M} = \bm{W}$ under a collocation assumption, we utilize $\bm{W}$ for the remainder of this work to emphasize that the mass matrix is diagonal and related to the quadrature weights $w_i$.  The treatment of non-diagonal mass matrices is covered in \cite{chan2017discretely, chan2018discretely}.

@@

Let $f(x), g(x)$ denote two functions, and let $\bm{f}, \bm{g}$ denote the values of $f,g$ at interior nodal points.  We also define vectors $\bm{f}_N, \bm{g}_N$ denoting the values of $f,g$ at both interior and boundary points 
\begin{equation}
\bm{f}_N = \begin{bmatrix}
f\LRp{x_1}\\
\vdots\\
f\LRp{x_{N+1}}\\
f(-1)\\
f(1)
\end{bmatrix} = \begin{bmatrix}
\bm{f} \\
\bm{f}_f
\end{bmatrix}, \qquad
\bm{g}_N = \begin{bmatrix}
g\LRp{x_1}\\
\vdots\\
g\LRp{x_{N+1}}\\
g(-1)\\
g(1)
\end{bmatrix} = \begin{bmatrix}
\bm{g} \\
\bm{g}_f
\end{bmatrix}.
\label{eq:fg}
\end{equation}
Then, a polynomial approximation to $f\pd{g}{x}$ can be computed using $\bm{Q}_N$.  Let $\bm{u}$ denote the nodal values of the polynomial $u(x) \approx f\pd{g}{x}$.  These coefficients are computed via
\begin{equation}
\bm{W}\bm{u} = \begin{bmatrix}
\bm{I}\\
\bm{V}_f
\end{bmatrix}^T \diag{\bm{f}_N} \bm{Q}_N \bm{g}_N.
\label{eq:qn}
\end{equation}
The expression (\ref{eq:qn}) can be rewritten in ``strong'' form as follows
\begin{align*}
\bm{u} = \bm{W}^{-1}\begin{bmatrix}
\bm{I}\\
\bm{V}_f
\end{bmatrix}^T& \diag{\bm{f}_N} \bm{Q}_N \bm{g}_N \\
= \diag{\bm{f}}\bm{D}\bm{g} &+ \frac{1}{2}\diag{\bm{f}} \bm{W}^{-1} \bm{V}_f^T \bm{B}\LRp{\bm{g}_f - \bm{V}_f\bm{g}} \\
&+ \frac{1}{2}\bm{W}^{-1} \bm{V}_f^T\bm{B}\diag{\bm{f}_f} \LRp{\bm{g}_f - \bm{V}_f\bm{g}},
\end{align*}
where we have used the fact that diagonal matrices commute to simplify expressions.  The decoupled SBP operator $\bm{Q}_N$ can thus be interpreted as adding boundary corrections to the GSBP operator $\bm{D}$ in a skew-symmetric fashion.

@@

Finally, we define $\bm{F}_S$ as the matrix of evaluations of the two-point flux $\bm{f}_S$ at combinations of values of $\tilde{\bm{u}}$
\[
\LRp{\bm{F}_S}_{ij} \myeq \bm{f}_S\LRp{\tilde{\bm{u}}_i,\tilde{\bm{u}}_j}, \quad 1\leq i,j \leq 2(N+3).
\]
Note that, due to the consistency of $\bm{f}_S$, the diagonal of $\bm{F}_S$ reduces to flux evaluations
\begin{equation}
\label{eq:diagFS}
\LRp{\bm{F}_S}_{ii} = \bm{f}_S\LRp{\tilde{\bm{u}}_i,\tilde{\bm{u}}_i} = \bm{f}\LRp{\tilde{\bm{u}}_i}.  
\end{equation}
We can now construct a semi-discretely entropy conservative formulation based on decoupled SBP operators:
\begin{theorem}
\label{thm:consentropy}
Let $\bm{Q}_h$ be defined by (\ref{eq:decoupledsbp2elem}), and let $\bm{u}_h$ denote the two-element solution of the following formulation:
\begin{gather}
\label{eq:form}
\bm{W}_h \td{}{t} \bm{u}_h + \bm{I}_h^T\LRp{\bm{Q}_h \circ \bm{F}_S}\bm{1} = 0.
\end{gather}
Then, $\bm{u}_h$ satisfies a semi-discrete conservation of entropy
\begin{gather*}
\bm{1}^T\bm{W}_h\td{S(\bm{u}_h)}{t} + \bm{1}^T\bm{B}\LRp{\bm{v}_f^T\bm{f}(\tilde{\bm{u}}_f) - \psi\LRp{\tilde{\bm{u}}_f}} = 0\\
\bm{v}_f \myeq \begin{bmatrix}
\bm{t}_L^T \bm{v}\LRp{\bm{u}_N^1}\\
\bm{t}_R^T \bm{v}\LRp{\bm{u}_N^2}
\end{bmatrix}, \qquad \tilde{\bm{u}}_f \myeq \bm{u}(\bm{v}_f).
\end{gather*}
%Let $\circ$ denote the Hadamard product, and let $\tilde{\bm{u}}_f^+$ denote values of the entropy-projected conservative variables on neighboring elements.  Then, the formulation
%\begin{equation}
%\label{eq:form}
%\bm{W}\td{\bm{u}_N}{t} + 2 \begin{bmatrix}
%\bm{I}\\
%\bm{V}_f
%\end{bmatrix}^T\LRp{\bm{Q}_N \circ \bm{F}_S}\bm{1} + \bm{V}_f^T \bm{B} \LRp{\bm{f}_S\LRp{\tilde{\bm{u}}_f^+,\tilde{\bm{u}}_f} - \bm{f}(\bm{u})} = 0
%\end{equation}
%satisfies the quadrature form of the semi-discrete conservation of entropy
%\[
%\bm{1}^T\bm{W}\td{S(\bm{u}_N)}{t} + \bm{1}^T\bm{B} \LRp{\bm{v}_f^T\bm{f}(\tilde{\bm{u}}_f) - \psi_i\LRp{\tilde{\bm{u}}_f}} = 0.
%\]
\end{theorem}
\begin{proof}
The proof results from testing with $\bm{v}_h^T$.  Since $\bm{v}(\bm{u}) = \pd{S(\bm{u})}{\bm{u}}$ and $\bm{W}_h$ is diagonal, the time term yields
\[
\bm{v}_h^T\td{}{t}\bm{W}_h\bm{u}_h = \bm{1}^T \td{}{t}\bm{W}_h\diag{\bm{v}_h}\bm{u}_h = \bm{1}^T\td{}{t}\bm{W}_h S(\bm{u}_h).
\]
The spatial term can be manipulated as follows
\begin{align*}
2\bm{v}_h^T \bm{I}_h^T\LRp{\bm{Q}_h \circ \bm{F}_S}\bm{1} &= 2\LRp{\bm{I}_h\bm{v}_h}^T \LRp{\LRp{\bm{S}_h \circ \bm{F}_S}\bm{1} + \frac{1}{2}\LRp{\bm{B}_h \circ \bm{F}_S}\bm{1}}\\
&= \tilde{\bm{v}}^T\LRp{\bm{B}_h \circ \bm{F}_S}\bm{1} + \tilde{\bm{v}}^T \LRp{\bm{S}_h \circ \bm{F}_S}\bm{1} - \bm{1}^T \LRp{\bm{S}_h \circ \bm{F}_S}\tilde{\bm{v}},
\end{align*}
where we have used the skew-symmetry of $\bm{S}_h$ in the last step.  The boundary term reduces to 
\[
\tilde{\bm{v}}^T\LRp{\bm{B}_h \circ \bm{F}_S}\bm{1} = \bm{1}^T\bm{B}_h\tilde{\bm{v}}^T\bm{f}\LRp{\tilde{\bm{u}}}= \bm{1}^T\bm{B}\tilde{\bm{v}}_f^T\bm{f}\LRp{\tilde{\bm{u}_f}},
\]
where we have used (\ref{eq:diagFS}) and the fact that $\bm{B}_h$ is diagonal.  Here, $\tilde{\bm{v}}^T\bm{f}\LRp{\tilde{\bm{u}}}$, $\tilde{\bm{v}}_f^T\bm{f}\LRp{\tilde{\bm{u}}_f}$ denote vectors whose entries are the vector inner products of components of $\tilde{\bm{v}}$ and $\bm{f}\LRp{\tilde{\bm{u}}}$ at volume and face points, respectively.

@@

It is possible to convert the semi-discrete entropy equality in Theorem~\ref{thm:consentropy} to a semi-discrete entropy inequality by adding appropriate interface dissipation terms, such as Lax-Friedrichs or matrix dissipation \cite{winters2017uniquely}.  We note that these terms must be computed in terms of $\tilde{\bm{u}}_f$ in order to ensure a discrete dissipation of entropy \cite{chen2017entropy, chan2017discretely}.

@@

\section{Numerical results}
\label{sec:3}
The compressible Euler equations in $d$ dimensions are given as follows:
\begin{align*}
\pd{\rho}{t} &+ \sum_{j=1}^d \pd{\LRp{\rho \bm{u}_j}}{x_j} = 0,\\
\pd{\rho \bm{u}_i}{t} &+ \sum_{j=1}^d \pd{\LRp{\rho \bm{u}_i\bm{u}_j + p\delta_{ij} }}{x_j} = 0, \qquad i = 1,\ldots,d\\
\pd{E}{t} &+ \sum_{j=1}^d \pd{\LRp{\bm{u}_j(E+p)}}{x_j} = 0.\nonumber
\end{align*}
Here, $\rho$ is density, $\bm{u} = \LRp{\bm{u}_1, \ldots,\bm{u}_d}$ is the vector of velocities, and $E$ is the total energy.  The pressure $p$ and specific internal energy $\rho e$ are given by 
\[
p = (\gamma-1)\LRp{E - \frac{1}{2}\rho \LRb{\bm{u}}^2}, \qquad \rho e = E - \frac{1}{2}\rho \LRb{\bm{u}}^2, \qquad \LRb{\bm{u}}^2 = \LRp{\sum_{j=1}^d \bm{u}_j^2}.
\]

@@

v_{1+ i}&= \frac{\rho \bm{u}_i}{\rho e}, \qquad i = 1,\ldots, d,\\

@@

 \rho \bm{u}_i &= (\rho e) v_{1+i}, \qquad i = 1,\ldots,d\\

@@

@@

&f^1_{1,S}(\bm{u}_L,\bm{u}_R) = \avg{\rho}^{\log} \avg{u},& &f^1_{2,S}(\bm{u}_L,\bm{u}_R) = \avg{\rho}^{\log} \avg{v},&\\
&f^2_{1,S}(\bm{u}_L,\bm{u}_R) = f^1_{1,S} \avg{u} + p_{\rm avg},&  &f^2_{2,S}(\bm{u}_L,\bm{u}_R) = f^1_{2,S} \avg{u},&\nonumber\\
&f^3_{1,S}(\bm{u}_L,\bm{u}_R) = f^2_{2,S},& &f^3_{2,S}(\bm{u}_L,\bm{u}_R) = f^1_{2,S} \avg{v} + p_{\rm avg},&\nonumber\\
&f^4_{1,S}(\bm{u}_L,\bm{u}_R) = \LRp{E_{\rm avg} + p_{\rm avg}}\avg{u},& &f^4_{2,S}(\bm{u}_L,\bm{u}_R) = \LRp{E_{\rm avg} + p_{\rm avg} }\avg{v},& \nonumber

@@

where we have defined the auxiliary quantities 

@@

p_{\rm avg} = \frac{\avg{\rho}}{2\avg{\beta}}, \qquad E_{\rm avg} = \frac{\avg{\rho}^{\log}}{2\avg{\beta}^{\log}\LRp{\gamma -1}}   + \frac{\nor{\bm{u}}^2_{\rm avg}}{2}, \label{eq:fluxaux} \\
 \nor{\bm{u}}^2_{\rm avg} = 2(\avg{u}^2 + \avg{v}^2) - \LRp{\avg{u^2} +\avg{v^2}} \nonumber.  

@@

\avg{\rho}^{\log}\avg{u}\\
\avg{\rho}^{\log}\avg{u}^2 + p_{\rm avg}\\
\avg{\rho}^{\log}\avg{u}\avg{v}\\
\avg{\rho}^{\log}\avg{u}\avg{w}\\
(E_{\rm avg}+ p_{\rm avg})\avg{u}\\

@@

\avg{\rho}^{\log}\avg{v}\\
\avg{\rho}^{\log}\avg{u}\avg{v}\\
\avg{\rho}^{\log}\avg{v}^2 + p_{\rm avg}\\
\avg{\rho}^{\log}\avg{v}\avg{w}\\
(E_{\rm avg}+ p_{\rm avg})\avg{v}\\

@@

\avg{\rho}^{\log}\avg{w}\\
\avg{\rho}^{\log}\avg{u}\avg{w}\\
\avg{\rho}^{\log}\avg{v}\avg{w}\\
\avg{\rho}^{\log}\avg{w}^2 + p_{\rm avg}\\
(E_{\rm avg}+ p_{\rm avg})\avg{w}\\

@@

with auxiliary quantities 
p_{\rm avg} &= \frac{\avg{\rho}}{2\avg{\beta}}, \qquad E_{\rm avg} = \frac{\avg{\rho}^{\log}}{2(\gamma-1)\avg{\beta}^{\log}} + \frac{1}{2}\avg{\rho}^{\log}\nor{\bm{u}}^2_{\rm avg}, \\
\nor{\bm{u}}^2_{\rm avg} &= 2(\avg{u}^2 + \avg{v}^2 + \avg{w}^2) - \LRp{\avg{u^2} +\avg{v^2} + \avg{w^2}}.\nonumber
